Does an earthquake “know” how big it’s going to be? A neural-net aided study

### 2024-02-22 article.tex
Edit introduction.

\title[Running Title]{Using past seismicity to predict the magnitude of future earthquakes \newline or \newline Does an earthquake “know” how big it’s going to be? A neural-net aided study}

@@

Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{geller_earthquakes_1997, bernard_earthquake_1999} (cite nature debate on earthquake prediction? EQs cannot be predicted Kagen?). Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes. This includes various characteristics of spatial and temporal clustering (cite Omori, Kagan, benzion \& Zaliapin, many others, ), which can be incorporated into stochastic prediction schemes (ETAS papers, Hardebeck etal Ann Rev 2024, Meade on NN aftershock location, some CFS paper). These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence. These spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability (cite and read Rathbun, 1993; Schoenberg, 2004). That is, while seismic history is used to inform the location and rate of future earthquakes, i.e.~by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events is determined by drawing values from the Gutenberg-Richter (GR) distribution or its variants (Gutenberg and Richter, 1944;  a kagen tapered GR, some spatiotemporal variation of GR, Ogata's magnitude prediction variational GR paper as well?). Importantly, this distribution does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events. %Such stochastic models, coupled with the separability assumption, imply that earthquake magnitude emergence is a stochastic process, independent of seismic history and spatial distribution, suggesting

\neri{This entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere?}
This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

While it is intelectually compelling, the assumption of separability is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field, which in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability is not a priori justified, and its validity should be tested empirically. Information about an earthquake magnitude exists before its initiation, but may be inaccessible using our current experimental or analytic tools. In this work we directly address this question, and ask whether some information about the magnitude of a given earthquake can be extracted from regional seismic history.

\neri{(short literature review)}

Here we address this question directly by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and log term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

This contribution has two important consequences. From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable. Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction. This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.

% paradigm of earthquake  This both   . We assert that the standard hypocenter catalog alone contains sufficient information to predict future earthquake magnitudes with better-than-random accuracy. This contradicts the common view that such catalogs represent a stochastic process of earthquake magnitude occurrence (cite, perhaps Beroza etal Nat Comm 2021? their claim is that you need a ML generated catalog better predict earthquake properties). The presented results demonstrate that incorporating spatial features and seismic history into magnitude forecasting yields an information gain, thereby answering the question posed above.

We construct an encoder-decoder neural network (NN). Our model uses as input a hypocenter catalog to construct a latent representation of the seismic state at a specific timestamp. Then, the latent representation is passed, together with  a space-time coordinate query of a future earthquake, into a decoder that outputs a parameterized probability density function (pdf) of the earthquake. See \ref{fig:architecture} and caption for detailed description of the neural network's architecture. We optimize the model parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e. the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set. We use the \yohai{kamawarakskakaskasasi} (cite, refer to methods/SI) parameterized family of probability distributions, which is defined by \yohai{N} parameters. This family smoothly can interpolate between exponential distribution, like GR, and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We train the model on four distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al. (cite) and SCSN (cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...] (refer to Supplementary Figure and Table for details). While all four catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the pdf.

@@

\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/intro_fig.pdf}
	\caption{
 Architecture of the magnitude prediction neural network model. \textbf{a} The input catalog is encoded with a set of predefined feature functions at multiple timestamps. These features are then used as an input to an LSTM network to create an encoded latent representation of the catalog. The latent representation is combined with a space-time query and fed into a fully connected neural network for decoding the data. The decoding network outputs a set of parameters defining a probability density function for each time-space query coordinates. \textbf{b} The pdfs from 100 random samples from the test set of the Hauksson catalog, color coded by the magnitude of the earthquake that has occurred in the corresponding query coordinates. The black dashed line shows Gutenberg Richter distribution, fitted for the train set.
}
\label{fig:architecture}
\end{figure}

@@

>>>>

\title[Running Title]{Does an earthquake “know” how big it’s going to be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes

@@

Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}. Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes. This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}. These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence. Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}. That is, while seismic history is used to inform the location and rate of future earthquakes, i.e. by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}. Importantly, this distribution does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.

% This might be something to remove to shorten
% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

While assuming that an earthquake's magnitude is independent of the seismic history is popular sue to its simplicity and its ease of use in earthquake modeling, it is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field. These in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability, i.e. its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools. Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding. 
\newline
In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history. This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

This contribution has two important consequences. From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}. Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction. This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.

\neri{Is a literature review nescessary here or is it ingested enough in the text itself? I'm pretty satisfied with how the paragraphs up to now are written.}\newline
\neri{The entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere in the literature review? can be seem as comment in source code.}
% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

We construct a neural network (NN) model that encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place. The constructed NN uses a long-short term memory (LSTM) architecture to model the seismic history up to a given time. This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized probability density function (PDF) of the possible magnitudes at the quarried space- time coordinates. This procedure is depicted in Fig. \ref{fig:intro_fig}a depicts the task as described here. A detailed description of the model's architecture is given in the methods section. We optimize the model parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e. the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set. We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain. The final distribution used is defined by 5 parameters. This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We train the model on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al. (cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...]. While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the PDF.

@@

\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/intro_fig.pdf}
	\caption{
 Caption.
}
\label{fig:intro_fig}
\end{figure}

@@

Focusing on existing events within the catalog due to our model's requirement for labeled data, our analysis utilizes the space-time coordinates of earthquakes as input. This necessitates excluding queries about arbitrary spatiotemporal points during training and evaluation, as these processes require knowledge of the true magnitude for each event at its specific location and time. Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location. Figure \ref{fig:architecture}b showcases PDFs produced by our model, $p_{\textbf{x}_i, t_i}(m)$, for 100 randomly sampled event times and locations from the test set. Included as a baseline is the stationary GR predictor trained on the training set, representing its potential operational use (e.g., Tsuruoka et al., 2012). A clear trend emerges: PDFs predicted for higher-magnitude events (warm colors) exhibit a greater skew towards higher magnitudes compared to those for lower-magnitude events (cooler colors). Figure \ref{fig:output_pdfs}a further highlights this trend by displaying a selection of model-generated PDFs alongside the corresponding likelihood of retrieving the true magnitude at the time and location each PDF was estimated for, denoted by a [color] circle.

Figure \ref{fig:output_pdfs}b focuses solely on the likelihood scores for retrieving the true magnitude, omitting the underlying PDFs used for their estimation. Additionally, it compares these scores against those generated by various benchmark models. Notably, our model demonstrates an advantage for most events exceeding magnitude $\sim 3$, while slightly under performing for lower-magnitude events. However, the overall mean log-likelihood (LL) score defined in \ref{eq:loss_function} (evaluated using the model-specific $p_{\textbf{x}_i, t_i}(\cdot)$ for each event) reveals a consistent advantage for our model across all datasets examined as summarized in \ref{tab:mean_ll}, where our model consistently obtains the lowest minus mean LL score in each column.

The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
\begin{equation}
    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(model)}(m_i)} - \log{p^{(GR)}(m_i)}
    \label{eq:information_gain}
\end{equation}
This equation measures the number of bits of information gained by using our model compared to the GR benchmark. Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage. \ref{fig:likelihood_over_time}a shows the Hauksson data test set event magnitudes at their occurrence times, colored according to the information gain. This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as a gray dashed line in the figure. The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods. Notably, for the Hauksson data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes. Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4$ bits, respectively, though these events are typically considered completely unpredictable. This observation leads to a significant corollary: at least some portion of the information gain likely originates from a physical precursory signal. This point will be further elaborated upon in the discussion section.
\newline
The observed increase in information gain persists throughout the entire test set for all regions examined in this study. This includes regions previously analyzed by (Ogata et al. SRL 2018), where no information gain was detected using the method employed at the time. Further analysis of \ref{fig:likelihood_over_time} applied to other regions is presented in the supplementary material.
\newline
While the information gain observed in pre- and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)? To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:

\begin{equation}
    p \left( m \vert m_i > \Tilde{m} \right) = \frac{p_{\pmb{x}_i, t_i}(m)} {\int_{\Tilde{m}}^{\infty} p(m') ,dm'}
    \label{eq:conditioned_likelihood}
\end{equation}

For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events. The resulting temporal incompleteness over time is displayed in \ref{fig:likelihood_over_time}b.
Crucially, \ref{fig:likelihood_over_time}b also presents the magnitudes of the Ridgecrest sequence at their origin times, colored by the information gain of our model over the GR benchmark, both conditioned on values above $m_c(t)$ as in \ref{eq:conditioned_likelihood}. This figure clearly demonstrates that even after accounting for temporal incompleteness, the information gain achieved by our model remains an increasing function throughout the test set. The mean score of the test set for the conditioned results are shown in [[]] which displays an outpreformance for our model with this metric as well. This finding underscores the ability of our model to capture true physical features beyond mere measurement artifacts.

[TODO: should I keep of remove the metrics for the causality defying metrics (present events) that beat the model in the ROC? this affects how the rest of this chapter will be written]
